A Mechanised Library for Generic (Co)Programming

### 2024-02-23 paper/paper.tex
Recursive coalgebras

We provide a custom induction scheme for initial algebras:
\begin{minted}{Coq}
Lemma LFix_ind [P : LFix -> Prop]
  : (forall x : App F LFix,
        (forall e : Pos (shape x), P (cont x e)) ->
        P (LFix_in x))
    -> forall l : LFix, P l.
\end{minted}
We define the equality of initial algebras of container extensions recursively,
by traversing the structure and equating the shapes and positions. This
definition is equivalent to taking the fixpoint of \mintinline{Coq}{AppR} in
Section~\ref{sec:TODO}, where the shapes must be equivalent and, for equal
positions, the contents must be related by \mintinline{Coq}{LFixR}.
\begin{minted}{Coq}
Fixpoint LFixR (x y : LFix) : Prop :=
  let f_x := LFix_out x in
  let f_y := LFix_out y in
  shape f_x =e shape f_y /\
    (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
\end{minted}
From such initial algebras, we define \emph{catamorphisms} by recursively
traversing their structure. As usual, catamorphisms take an
\mintinline{Coq}{F}-algebra with carrier \mintinline{Coq}{A}, and produces a
function from \mintinline{Coq}{LFix F} to \mintinline{Coq}{A}.
\begin{minted}{Coq}
Definition cata_f `{equiv A} (g : Alg F A) : LFix -> A
:= fix f (x : LFix) :=
  match x with
  | LFix_in ax =>
      let (sx, kx) := ax in
      g (MkCont sx (fun e => f (kx e)))
  end.
\end{minted}
It is easy to show that this function is a proper morphism, and we use this
fact to encode them as a higher-order morphism, with the following type:
\begin{minted}{Coq}
Definition cata `{eA : equiv A} : Alg F A ~> LFix ~> A := ...
\end{minted}
Finally, we prove that catamorphisms satisfy their expected universal law (i.e.
they are the unique solution to the equation \mintinline{Coq}{f =e g \o fmap f \o l_out}).
The proof is by induction on the input to \mintinline{Coq}{f} and
\mintinline{Coq}{cata g}.
\begin{minted}{Coq}
Lemma cata_univ `{eA : equiv A} (g : Alg F A) (f : LFix ~> A)
    : f =e cata g <-> f =e g \o fmap f \o l_out.
\end{minted}

@@

  Lemma ana_univ `{eA : equiv A} (h : Coalg F A) (f : A ~> GFix)
    : f =e ana h <-> f =e g_in \o fmap f \o h.
\end{minted}
This definition, however, prevents us from composing catamorphisms and
anamorphisms. To address this issue, we consider another alternative definition
for initial $F$-algebras. In particular, we show that the terminal coalgebras
for $F$, coupled with a proof that they are finite are also initial
$F$-algebras:
\begin{minted}{Coq}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{minted}
Using this predicate, we can define the notion of \emph{recursive coalgebras}
as those that build finite trees. Recursive coalgebras are implemented in Coq
recursively on the proof that they produce finite trees.  Such proofs will be
erased when extracting code to OCaml. This is standard practice when building
general recursive functions in Coq\footnote{\dcas{This technique had a name in
the community, but I cannot remember. Something predicates}}.
\begin{minted}{Coq}
Notation rana_f__ h :=
  (fix f x H :=
    let hx := h x in
    LFix_in (MkCont (shape hx) (fun e => f (cont hx e) (RecF_inv H e)))).

Definition rana `{eA : equiv A} : RCoalg A ~> A ~> LFix F :=...

>>>>

We provide a custom induction scheme for initial algebras:
\begin{minted}{Coq}
Lemma LFix_ind [P : LFix -> Prop]
  : (forall x : App F LFix,
        (forall e : Pos (shape x), P (cont x e)) ->
        P (LFix_in x))
    -> forall l : LFix, P l.
\end{minted}
We define the equality of initial algebras of container extensions recursively,
by traversing the structure and equating the shapes and positions. This
definition is equivalent to taking the fixpoint of \mintinline{Coq}{AppR} in
Section~\ref{sec:TODO}, where the shapes must be equivalent and, for equal
positions, the contents must be related by \mintinline{Coq}{LFixR}.
\begin{minted}{Coq}
Fixpoint LFixR (x y : LFix) : Prop :=
  let f_x := LFix_out x in
  let f_y := LFix_out y in
  shape f_x =e shape f_y /\
    (forall e1 e2, val e1 = val e2 -> LFixR (cont f_x e1) (cont f_y e2)).
\end{minted}
From such initial algebras, we define \emph{catamorphisms} by recursively
traversing their structure. As usual, a catamorphism takes an
\mintinline{Coq}{F}-algebra with carrier \mintinline{Coq}{A}, and produces a
function from \mintinline{Coq}{LFix F} to \mintinline{Coq}{A}.
\begin{minted}{Coq}
Definition cata_f `{equiv A} (g : Alg F A) : LFix -> A
:= fix f (x : LFix) :=
  match x with
  | LFix_in ax =>
      let (sx, kx) := ax in
      g (MkCont sx (fun e => f (kx e)))
  end.
\end{minted}
It is easy to show that this function is a proper morphism, and we use this
fact to encode them as a higher-order morphism, with the following type:
\begin{minted}{Coq}
Definition cata `{eA : equiv A} : Alg F A ~> LFix ~> A := ...
\end{minted}
Finally, we prove that catamorphisms satisfy their expected universal law (i.e.
they are the unique solution to the equation \mintinline{Coq}{f =e g \o fmap f \o l_out}).
The proof is by induction on the input to \mintinline{Coq}{f} and
\mintinline{Coq}{cata g}.
\begin{minted}{Coq}
Lemma cata_univ `{eA : equiv A} (g : Alg F A) (f : LFix ~> A)
    : f =e cata g <-> f =e g \o fmap f \o l_out.
\end{minted}

@@

  Lemma ana_univ `{eA : equiv A} (h : Coalg F A) (f : A ~> GFix)
    : f =e ana h <-> f =e g_in \o fmap f \o h.
\end{minted}
This definition, however, prevents us from composing catamorphisms and
anamorphisms. To address this issue, we consider another alternative definition
for initial $F$-algebras. In particular, we show that the terminal coalgebras
for $F$, coupled with a proof that they are finite are also initial
$F$-algebras:
\begin{minted}{Coq}
Inductive FinF : GFix F -> Prop :=
| FinF_fold (x : GFix F) : (forall e, FinF (cont (g_out x) e)) -> FinF x.
\end{minted}
Catamorphisms from this initial algebra are denoted by \mintinline{Coq}{ccata},
and they can be used to define the isomorphism between 
\mintinline{Coq}{{x : GFix F | FinF x}} and
\mintinline{Coq}{LFix F}. In particular: 
\begin{minted}{Coq}
Lemma cata_ccata `{equiv A} (f : Alg F A) : cata f \o ccata l_in =e ccata f.
Lemma ccata_cata `{equiv A} (f : Alg F A) : ccata f \o cata lg_in =e cata f.
\end{minted}
Using this new initial algebra, we can compose catamorphisms and anamorphisms,
\emph{as long as the anamorphisms produce finite trees}. However, if
anamorphisms produce finite trees, then we can also encode in Coq the notion of
\emph{recursive anamorphisms} and \emph{recursive coalgebras}~\cite{TODO}.
Recursive coalgebras are coalgebras whose anamorphism produces finite trees.
The following predicate captures this fact. Note how all the predicate states
is that \mintinline{Coq}{h} can only be applied finitely many times:
\begin{minted}{Coq}
Inductive RecF (h : Coalg F A) : A -> Prop :=
| RecF_fold x : (forall e, RecF h (cont (h x) e)) -> RecF h x.
\end{minted}
We show how this predicate is equivalent to the anamorphism of such coalgebras
producing finite trees:
\begin{minted}{Coq}
Lemma ana_rec_term `{equiv A} (h : Coalg F A)
  : (forall x, FinF (ana h x)) <-> (forall x, RecF h x).
\end{minted}
Recursive anamorphisms are implemented in Coq recursively on the proof that
they can only be applied finitely many times (\mintinline{Coq}{RecF}).  Note
that such proofs will be erased when extracting code to OCaml. This is standard
practice when building general recursive functions in Coq\footnote{\dcas{This
technique had a name in the community, but I cannot remember. Something
predicates}}. The core definitions are the following:
\begin{minted}{Coq}
Lemma RecF_inv (h : Coalg F A) x : RecF h x -> forall e, RecF h (cont (h x) e).

Definition rana_f__ `{equiv A} (c : Coalg F A)
  : forall x : A, RecF c x -> LFix F
  := fix f (x : A) (H : RecF c x) :=
       let hx := c x in
       LFix_in (MkCont (shape hx) (fun e => f (cont hx e) (RecF_inv H e))).
\end{minted}
Here, \mintinline{Coq}{RecF_inv} provides the structurally smaller element for
the Coq recursive definition. We then couple coalgebras with a proof that they
are recursive, and add a coercion from recursive coalgebras to coalgebras.
\begin{minted}{Coq}
Definition RCoalg F A = {coalg :> Coalg F A; recP : forall x, RecF coalg x}.
\end{minted}
Similarly to the rest of the definitions, we prove that recursive anamorphisms
are respectful, and use this fact to build them as higher order proper
morphisms, and prove its universal law, i.e.\ that recursive anamorphisms are
the unique solution to the equation 
\mintinline{Coq}{f =e l_in \o fmap f \o h},
where \mintinline{Coq}{h} is a recursive coalgebra.
\begin{minted}{Coq}
Definition rana `{eA : equiv A} : RCoalg A ~> A ~> LFix F :=...
